---
title: "DATA INGESTION"
author: "Dajana Ramirez Andrade"
date: "`r Sys.Date()`"
output: html_document
---

Loading the the library that I will be using for data ingestion. 
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
```

Reading in only the data set that is currently available to me. (From 02/15/2025) 
```{r}
tuition_data <- read_csv("Tuition_Assistance_20250215.csv")
```
Checking for missing data in my data set.
```{r}
colSums(is.na(tuition_data))
```
Since there is no missing data I will just mutate the the characters for there to be no whitespace.
```{r}
tuition_clean <- tuition_data %>%
  clean_names() %>%                     
  mutate(across(where(is.character), str_trim))  
```

Looking at my cleaned data set to make sure evrything is ready for analysis.
```{r}

glimpse(tuition_clean)
summary(tuition_clean)

```
Everything looks alright and in there place, no inconsistencies. 


Making sure some one time courses aren't treated as a degree.
```{r}
tuition_clean %>%
  count(degree) %>%
  ggplot(aes(x = reorder(degree, -n), y = n, fill = degree)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Schools by Degree Type",
       x = "Degree Type",
       y = "Count",
       fill = "Degree Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Looks like they aren't being treated as a degree.


Dashboard: https://public.tableau.com/views/PHASE1_17439510545370/PHASE1DASHBOARD?:language=en-US&:sid=&:redirect=auth&:display_count=n&:origin=viz_share_link

I have loaded and cleaned my data set to start working on my analysis, along with that Ive explored my data in tableu and made a dashboard. My findings to the following questions were:

1. Which departments receive the most tuition assistance in dollars?

The police and fire rescue services. 

2. Which majors receive the most tuition assistance?

That would be non-majors such as certificates, along with Business and Criminal Justice.

3. Which schools or universities are most attended by employees? 

University of Maryland and Montgomery College Rockville Campus. 

4. What is the distribution of costs across different degrees (Bachelor’s vs Master’s)?

There is more people completing a bachelors degree (almost 2 times bachelors), but the cost are almost the same, so masters are more expensive. 

5. Do certain departments show strong preferences for specific schools?

Theres no much to this, but it seems that the police and health services lean towards University of Maryland.



