---
title: "DATA INGESTION"
author: "Dajana Ramirez Andrade"
date: "`r Sys.Date()`"
output: html_document
---

Loading the the library that I will be using for data ingestion. 
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(viridis)

```

Reading in only the data set that is currently available to me. (From 02/15/2025) 
```{r}
tuition_data2025 <- read_csv("Tuition_Assistance_20250215.csv")
```
Checking for missing data in my data set.
```{r}
colSums(is.na(tuition_data2025))
```
Since the only missing data appears to be in course description, I will not be removing it since I won't be using this variable. I will just mutate the the characters for there to be no whitespace.
```{r}
tuition_clean2025 <- tuition_data2025 %>%
  clean_names() %>%                     
  mutate(across(where(is.character), str_trim))  
```

Looking at my cleaned data set to make sure everything is ready for analysis.
```{r}
glimpse(tuition_clean2025)
summary(tuition_clean2025)

```
Everything looks alright and in there place, no inconsistencies. 


Making sure some one time courses aren't treated as a degree.
```{r}
tuition_clean2025 %>%
  count(degree) %>%
  ggplot(aes(x = reorder(degree, -n), y = n, fill = degree)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Schools by Degree Type 2025",
       x = "Degree Type",
       y = "Count",
       fill = "Degree Type") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Looks like they aren't being treated as a degree; they have their own categories such as certificate, non-degree, and other.


Dashboard: https://public.tableau.com/views/EDA1DATA2025/PHASE1DASHBOARD?:language=en-US&:sid=&:redirect=auth&:display_count=n&:origin=viz_share_link

I have loaded and cleaned my data set to start working on my analysis, along with that I"ve explored my data in tableu and made a dashboard. My findings to the following questions were:

1. Which departments receive the most tuition assistance in dollars?

The police and fire rescue services. 

2. Which majors receive the most tuition assistance?

That would be non-majors such as certificates, along with Business and Criminal Justice.

3. Which schools or universities are most attended by employees? 

University of Maryland and Montgomery College Rockville Campus. 

4. What is the distribution of costs across different degrees (Bachelor’s vs Master’s)?

There is more people completing a bachelors degree (almost 2 times bachelors), but the cost are almost the same, so masters are more expensive. 

5. Do certain departments show strong preferences for specific schools?

There's no much to this, but it seems that the police and health services lean towards University of Maryland.


Update 04/14: I haven't been able to get official acess for the previous data set, but I have found one in Kaggle which I believe originates from Montgomery Data, so I will be using that one. 

Data Source: https://www.kaggle.com/datasets/sahirmaharajj/school-tution-dataset

First off I will be reading in the data set, following the same steps as I did for the first data set. 

```{r}
tuition_data2024 <- read_csv("Tuition_Assistance_20240525.csv")
```
I will be checking for missing data in the secondary data set.

```{r}
colSums(is.na(tuition_data2024))
```
As previously seen in our previous data set, the missing data set only appears in the variable Course Description, since I will not be using this variable I will leave this data set as is. I will only be removing the white space from the variables that are characters.

```{r}
tuition_clean2024 <- tuition_data2024 %>%
  clean_names() %>%                     
  mutate(across(where(is.character), str_trim))  
```


I will also be looking at my data to make sure everything is ready for analysis. 

```{r}
glimpse(tuition_clean2024)
summary(tuition_clean2024)
```
Everything is looking alright, I do notice that compared to our data from 2025, this data from 2024 seems to have a lower mean, just by 1.7 dollars. They have the same median, min, and max, but they have different 1st quartiles and 3rd quartiles.

I will also be making sure that one time courses aren't treated as a degree, just as I did for data set of the year 2025. 

```{r}
tuition_clean2024 %>%
  count(degree) %>%
  ggplot(aes(x = reorder(degree, -n), y = n, fill = degree)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d() + 
  labs(title = "Number of Schools by Degree Type 2024",
       x = "Degree Type",
       y = "Count",
       fill = "Degree Type") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

As we can see, they aren't being treated as a degree, but rather they have their own categories such as certificate, non-degree, and other.



Dashboard: https://public.tableau.com/views/EDA2DATA2024/2024Data?:language=en-US&:sid=&:redirect=auth&:display_count=n&:origin=viz_share_link

I've also made a dashboard for this data set, I found that:

1. Which departments receive the most tuition assistance in dollars?

The police and fire rescue services, just like the previous data set. 

2. Which majors receive the most tuition assistance?

Just like the other data set it is non-majors such as certificates, along with Business and Criminal Justice.

3. Which schools or universities are most attended by employees? 

University of Maryland and Montgomery College Rockville Campus, just as the previous dataset. 

4. What is the distribution of costs across different degrees (Bachelor’s vs Master’s)?

There is more people completing a bachelors degree (almost 2 times bachelors), but the cost are almost the same, so masters are more expensive. 

5. Do certain departments show strong preferences for specific schools?

There's no much to this, but it seems that the police and health services lean towards University of Maryland and Montgomery community college.
