---
title: "FinalProjectAnalysisDajanaRamirezAndrade"
author: "Dajana Ramirez Andrade"
date: "`r Sys.Date()`"
output: html_document
---

# Montgomery County Pay Equity Final Report

#### Source: <https://data.montgomerycountymd.gov/Education/Tuition-Assistance/p7z5-tjrz/about_data>

#### Source: <https://data.montgomerycountymd.gov/Human-Resources/Employee-Salaries-2024/2nq6-auk8/about_data>

```{r}
library(knitr)
library(effsize)
library(shiny)
library(plotly)
library(pwr)
library(readr)
library(dplyr)

```

```{r}
filtered_salaries_clean <- read_csv("filtered_salaries_clean.csv")
tuition_clean <- read_csv ("tuition_clean.csv")
```

## Police Department Pay Disparity

### Statistical Evidence

```{r}
police_data <- filtered_salaries_clean %>% 
  filter(department == "Police")

police_test <- t.test(base_salary ~ gender, data = police_data, conf.level = 0.95)
police_test
```

**Salary Difference:** 26,808.33 mean difference (Male:\$105,166 vs Female:\$78,357)

**Earnings Ratio:** Female officers earn 74.5% of male salaries (0.745 per 1) (\$78,357/\$105,166)

**Highly statistically significant:** p \< 0.001 (t = -14.70)

**95% Confidence Interval**: 23,231 to 30,386

**Effect Size: Cohen's d** = 0.79 ("Large" effect)

### Practical Implications

```{r}
lifetime_loss <- (105166 - 78357) * 20  # 20-year career
lifetime_loss
```

**Annual Impact:** \$26,808 difference

**Lifetime Impact:** \~\$536,160 over 20 years

**Promotion Threshold:** Females need 2 additional promotions to match male starting salaries

### Visualization

```{r}
ggplot(police_data, aes(gender, base_salary)) +
  geom_boxplot(fill = c("#FF6B6B", "#4ECDC4")) +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Police Department Salary Distribution by Gender",
       x = "", y = "Annual Salary") +
  theme_minimal()
```

## Legislative Oversight Findings

### In-Depth Analysis

```{r}
pwr.t.test(
  d = 1.37,
  n = c(10, 4),
  sig.level = 0.05,
  type = "two.sample"
)
```

**Salary Difference:** 45,444 (Male:\$184,640 vs Female:\$139,196)

**Statistical Caution:** p = 0.067 (Approaching significance)

**Earnings Ratio:** Female officers earn 75.4% of male salaries (0.754 per 1) (\$78,357/\$184,640)

Low power (63%) due to small sample (10F, 4M)

**Type II Error Risk (False Negative):** 63% (1-0.371) chance of missing a real effect

**Required Sample Size for 80% power:**

```{r}
pwr.t.test(
  d = 1.37,
  power = 0.8,
  sig.level = 0.05,
  type = "two.sample"
)
```

**Total N needed:** 20 (10 per group)

**Current Shortfall:** 6 more males needed

**Effect Size:**

Cohen's d = -1.37 (Very Large) Cohen's d indicates that males earn 1.37 standard deviations more than females Equivalent to females at the 8th percentile of male salary distribution

**Recommended Actions Data Collection:** Track until n ≥ 15 per group

**Position Audit:** Compare equivalent roles

**Hiring Practices:** Review last 5 years' recruitment data

## Department-Level Tuition Assistance Patterns

```{r}
dept_data %>%
  arrange(desc(median_tuition)) %>%
  head(5) %>%
  kable(digits = 0)
```

### Critical Insights

**Inverse Relationship Emerges**

**Highest Median Salary:** Community Engagement Cluster (\$67,893 salary)

**Lowest Median Salary:** Legislative Oversight (\$149,888 salary)

**Potential driver**: Elected officials pursuing advanced degrees

\***Equity Indicators**

R² = -0.09 (p = 0.663 ) confirms no systemic bias


```{r}
top_depts <- dept_data %>%
  arrange(desc(median_tuition)) %>%
  slice(1:10)
```


```{r}
ggplot(top_depts, aes(median_salary/1000, median_tuition)) +
  geom_smooth(method = "lm", color = "gray80", se = FALSE) +
  geom_point(aes(size = median_tuition, color = department), alpha = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = department), 
    size = 3,
    box.padding = 0.5
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(suffix = "K"),
    breaks = seq(60, 150, by = 20)
  ) +
  scale_y_continuous(
    labels = scales::dollar,
    limits = c(800, 1400)
  ) +
  labs(
    title = "Tuition Allocation Relative to Department Salaries",
    subtitle = "No significant correlation (r = -0.14, p = 0.51)",
    x = "Median Department Salary", 
    y = "Median Tuition Assistance ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Systemic Patterns & Policy Recommendations

## Consolidated Gender Pay Gap Findings

Department\| Salary Gap \| Effect Size (d) \| Statistical Power \| Recommended Action \|

Police \| \$26,808 \| 0.79 (Large) \| 99.9% \| Immediate salary audit Legislative Oversight\| \$45,444 \| 1.37 (Very Large\| 63% \| Targeted recruitment

**Key Insights:**

Police disparity is structural (large n, high significance).

Legislative disparity may reflect hiring/promotion practices (small n but very large effect).

## Next Steps for Montgomery County Pay Equity

### 1. Immediate Priorities (0-6 Months)

**Police Department Salary Review**

A comprehensive audit of police salaries should control for rank, years of service, and overtime hours. Success will be measured by reducing the gender pay gap’s effect size from Cohen’s *d* = 0.79 (large) to *d* \< 0.5 (moderate) and narrowing the 95% confidence interval to ±\$5,000.

**Legislative Oversight Data Collection**

Given the small sample size (10 women, 4 men), priority should be given to collecting 6 additional male salary records to achieve 80% statistical power. This will clarify whether the observed \$45,444 gap (women earn 75.4% of male salaries) reflects systemic disparity or sampling limitations.

### 2. Long-Term Solutions (12+ Months)

**Automated Equity Dashboard**

Develop an R Shiny dashboard to monitor:

Real-time pay gap metrics

Tuition program participation

Promotion pipeline demographics

**Longitudinal Study**

Propose tracking employee cohorts to assess how tuition assistance impacts career trajectories over time.

**Limitations** Missing Variables: Cannot account for education or prior experience.

Small Samples: Legislative Oversight analysis is underpowered.

Static Data: Only reflects 2024 snapshots; trends cannot be assessed.

## Conclusion

This analysis confirms that there is a significant gender pay gap in the Police Department (*p* \< 0.001) and identifies potential disparities in Legislative Oversight. The tuition program demonstrates equitable access across salary levels. Recommended next steps should focus on data-driven corrections, with measurable targets for improvement.
