---
title: "ProjectAnalysisDajanaRamirezAndrade"
author: "Dajana Ramirez Andrade"
date: "`r Sys.Date()`"
output: html_document
---

# Tuition Assistance Data Set and Employee Salaries Data Set From Montgomery County Analysis

#### Source: <https://data.montgomerycountymd.gov/Education/Tuition-Assistance/p7z5-tjrz/about_data>

#### Source:<https://data.montgomerycountymd.gov/Human-Resources/Employee-Salaries-2024/2nq6-auk8/about_data>

Loading in the libraries, that I will be using:

```{r}
library(dplyr)
library(tidyverse)
```

## Gender Pay Gap Analysis

Loading in the clean data set:

```{r}
filtered_salaries_clean <- read_csv("filtered_salaries_clean.csv")
tuition_clean <- read_csv ("tuition_clean.csv")
```

## Gender Pay Gap Analysis

I will now conduct t-tests to examine gender pay gaps in departments with high disparities (Police and Legislative Oversight).

```{r}
# Police Department analysis
police_data <- high_disparity_depts %>%
  filter(department == "Police")

# Summary statistics
police_summary <- police_data %>%
  group_by(gender) %>%
  summarise(
    n = n(),
    mean_salary = mean(base_salary),
    median_salary = median(base_salary),
    sd_salary = sd(base_salary)
  )

cat("\nPolice Department Salary Summary by Gender:\n")
print(police_summary)

# T-test
t_test_police <- t.test(base_salary ~ gender, data = police_data)

cat("\n### Police Department Gender Pay Gap T-Test Results\n")
print(t_test_police)

# Cohen's d
police_d <- (police_summary$mean_salary[police_summary$gender == "M"] - 
              police_summary$mean_salary[police_summary$gender == "F"]) /
            sqrt(((police_summary$sd_salary[police_summary$gender == "F"]^2 * 
                  (police_summary$n[police_summary$gender == "F"] - 1)) +
                 (police_summary$sd_salary[police_summary$gender == "M"]^2 * 
                  (police_summary$n[police_summary$gender == "M"] - 1))) /
                 (sum(police_summary$n) - 2))

cat("\nEffect Size (Cohen's d):", round(police_d, 3), "\n")
```

```{r}

# Legislative Oversight analysis
legislative_data <- high_disparity_depts %>%
  filter(department == "Legislative Oversight")

# Summary statistics
legislative_summary <- legislative_data %>%
  group_by(gender) %>%
  summarise(
    n = n(),
    mean_salary = mean(base_salary),
    median_salary = median(base_salary),
    sd_salary = sd(base_salary)
  )

cat("\nLegislative Oversight Salary Summary by Gender:\n")
print(legislative_summary)

# T-test
t_test_legislative <- t.test(base_salary ~ gender, data = legislative_data)

cat("\n### Legislative Oversight Gender Pay Gap T-Test Results\n")
print(t_test_legislative)

#Cohen's d
legislative_d <- (legislative_summary$mean_salary[legislative_summary$gender == "M"] - 
                   legislative_summary$mean_salary[legislative_summary$gender == "F"]) /
                 sqrt(((legislative_summary$sd_salary[legislative_summary$gender == "F"]^2 * 
                       (legislative_summary$n[legislative_summary$gender == "F"] - 1)) +
                      (legislative_summary$sd_salary[legislative_summary$gender == "M"]^2 * 
                       (legislative_summary$n[legislative_summary$gender == "M"] - 1))) /
                      (sum(legislative_summary$n) - 2))

cat("\nEffect Size (Cohen's d):", round(legislative_d, 3), "\n")
```

### Police Department Findings:

#### 1. Statistical Significance

Extremely significant pay gap (t = -14.70, p \< .001).

95% CI [- 30 , 386, − 30,386,−23,231] confirms male salaries are significantly higher.

#### 2. Effect Size

Cohen's d = 0.79 (large effect size)

Interpretation: Male officers earn 0.79 standard deviations more than females.

Practical significance: This represents a substantial, meaningful difference.

#### 3. Salary Comparison

Female mean: \$78,357 (87,499 median).

Male mean: \$105,166 (111,326 median).

Absolute difference: \$26,809 (34.2% higher for males).

### Legislative Oversight Findings

#### 1. Statistical Significance

Marginally non-significant (t = -2.29, p = .067).

95% CI [-95,374,95,374,4,485] includes zero but leans negative.

Small sample size reduces statistical power.

#### 2. Effect Size

Cohen's d = 1.37 (very large effect size).

Interpretation: Male salaries are 1.37 standard deviations higher than females.

Since it's a small sample (10 Females, 4 Males) it makes this estimate less stable.

#### 3. Salary Comparison

Female mean: \$139,196 (131,997.8 median).

Male mean: \$184,640 (194,304.1 median).

Absolute difference: \$45,444 (32.6% higher for males).

### Key Takeaways

#### Police Department

Clear evidence of gender pay disparity.

The difference is both statistically and practically significant.

#### Legislative Oversight

Large point estimate but limited by a small sample.

Potential pay gap warrants monitoring.

### Comparative Insights

Both departments show male salary advantages greater than 30%.

Effect sizes exceed typical thresholds for "large" effects.

Patterns suggest systemic rather than isolated issues.

## Calculate Pearson's r between median salary and tuition \$ per department

```{r}

# Calculating median salary by department
median_salaries <- filtered_salaries_clean %>%
  group_by(department) %>%
  summarise(median_salary = median(base_salary, na.rm = TRUE))

# Calculating median tuition by department
median_tuition <- tuition_clean %>%
  group_by(department) %>%
  summarise(median_tuition = median(cost, na.rm = TRUE))
```

```{r}
# Merging the data
dept_data <- inner_join(median_salaries, median_tuition, by = "department")
```

```{r}
# Calculating Pearson's r
cor_result <- cor.test(dept_data$median_salary, dept_data$median_tuition, 
                      method = "pearson")

cat("Pearson's r between median salary and median tuition per department:\n")
cat("r =", round(cor_result$estimate, 3), "\n")
cat("95% CI:", round(cor_result$conf.int, 3), "\n") 
cat("p-value:", format.pval(cor_result$p.value, digits = 3), "\n")


if(cor_result$p.value < 0.05) {
  direction <- ifelse(cor_result$estimate > 0, "positive", "negative")
  cat("\nInterpretation: Significant", direction, "correlation (p < 0.05)\n")
  cat("Departments with higher median salaries tend to have", 
      ifelse(direction == "positive", "higher", "lower"), 
      "median tuition costs.")
} else {
  cat("\nInterpretation: No significant correlation found.")

}


plot(dept_data$median_salary, dept_data$median_tuition,
     xlab = "Median Salary ($)", ylab = "Median Tuition Cost ($)",
     main = "Department-Level Correlation")
abline(lm(median_tuition ~ median_salary, data = dept_data), col = "red")
```

Pearson's r between median salary and median tuition per department: r = -0.09 95% CI: -0.461 0.308 p-value: 0.663 Interpretation: No significant correlation found.
